---
title: "Orienteering Course Exploratory Data Analysis"
author: "Mark Shoun"
date: "13 December 2019"
output: html_document
---

```{r setup, include=FALSE}
source("_libraries.R")
source("load/load_data.R")
```

# Data Overview

This is the first in a series of posts analyzing orienteering data. Orienteering is a sport that involves racing from point to point over unfamiliar terrain while using only a map and compass to navigate. The most common type of orienteering event is on foot (referred to as foot-O), but there are variants for mountain biking and cross-countrying skiing. Most foot-O courses consist of a start and finish separated by 7-20 intermediate control points which must be visited in a specified order. Competitors start running at staggered intervals, so that each person must navigate the course for themselves.

Orienteering events in the format described above are referred to as "classic" events. Non-classic events may have variant rules like mass starts; allowing competitors some choice over what order to visit control points; scoring based on number of control points visited, rather than elapsed time; significantly shorter or longer distances between control points than is customary; or other such variants. In these analyses, we focus on classic events for the sake of consistency and tractability.

The data for this analysis comes from results of events organized by the Quantico Orienteering Club, which is based in the greater Washington, DC metropolitan area. The club has posted results for `r nrow(event.df)` events from January 2012 through the present, of which `r length(unique(classic.leg.summary.df$event_date))` events included sufficiently detailed data for at least one classic course for analytical purposes.

# Course Summary Statistics

A typical classic orienteering event may have up to seven separate courses of varying length and difficulty. In the United States, orienteering courses tend to follow a standard naming scheme based on colors. The simplest and shortest course is White, followed by Yellow, Orange, Brown, Green, Red, and Blue. The Brown through Blue courses are generally considered "advanced" courses; these courses generally pose similar degrees of navigational difficulty and differ mostly on length and degree of physical difficulty.

In our data set, we have data on 975 separate courses. The table below shows a summary of the number of courses in the data set, as well as a brief summary of a course's typical features.

```{r courses_by_color, echo=FALSE}
classic.leg.summary.df %>% 
    mutate(Color=course_clean) %>% 
    group_by(Color) %>% 
    summarize(`Number of Courses` = n(),
              `Mean Length (km)` = round(mean(length) / 1000, 2),
              `Mean Climb (m)` = round(mean(climb)),
              `Mean Slope` = round(mean(climb / length), 3),
              `Mean Controls` = round(mean(control_points), 1),
              `Mean Leg Distance` = round(sum(length) / sum(control_points + 1))) %>%
    knitr::kable()
```

# Course Length

The plot below shows the distribution of course length by course color. The dot shows the mean course length, the heavy line shows the range from the 25th percentile to 75th percentile, and the lighter line shows the range from the 5th to the 95th percentile.

```{r course_length_plot, echo=FALSE}
course.length.summary.df <- classic.leg.summary.df %>% 
    group_by(course_clean) %>% 
    summarize(mu = mean(length),
              q05 = quantile(length, 0.05),
              q25 = quantile(length, 0.25),
              q75 = quantile(length, 0.75),
              q95 = quantile(length, 0.95))

ggplot(course.length.summary.df) +
    aes(x=course_clean) +
    geom_hline(yintercept=0) +
    geom_linerange(aes(ymin=q05, ymax=q95), alpha=0.7) +
    geom_linerange(aes(ymin=q25, ymax=q75), alpha=0.9, size=2) +
    geom_point(aes(y=mu), color="red",  size=1.5) +
    labs(x="Course Color", y="Course Distance (m)") +
    theme_bw()
```

We can see that, as expected, the mean course length generally increases with course difficulty, as does the mean course climb (a measure of the expected elevation gain a competitor will face on a near-optimal route). The least overlap between courses of consecutive colors is between Brown and Green: almost every Green course is longer than almost any Brown course.

Brown courses are slightly shorter than Orange courses on average, with significantly fewer control points. This is by design: Brown courses are more difficult to navigate than Orange courses (which is partially reflected in the greater average distance between control points), but are not intended to be more physically challenging than Orange courses.

## Per-Event Course Length Comparison

The comparison above is useful, but it relies on averages over many events. Some events are at smaller venues, which dictate shorter course lengths across the board. (In fact, some events omit Blue courses entirely when there is insufficient space.) Therefore, a cleaner comparison to see how often the Brown course at a given event is shorter than the Orange course at the same event. This controls for differences in course length due to venue, course designer proclivities, and other external factors. We can also perform this comparison for other pairs of courses besides Orange and Brown. The results of this comparison are shown in the table below.

```{r course_inversion_rate, echo=FALSE}
event.color.df <- classic.leg.summary.df %>% 
    dcast(event_date ~ course_clean, value.var="length")

inversion.rate.summary.df <- data.frame(
    easier = c("White", "Yellow", "Orange", "Brown", "Green", "Red", "Yellow", "Orange"),
    harder = c("Yellow", "Orange", "Brown", "Green", "Red", "Blue", "Brown", "Green"),
    stringsAsFactors = FALSE
)

inversions <- mapply(
    function(easy, hard) {
        comparison.df <- event.color.df[!(is.na(event.color.df[[easy]]) 
                                          | is.na(event.color.df[[hard]])), ]
        c(nrow(comparison.df), sum(comparison.df[[easy]] > comparison.df[[hard]]))
    }, inversion.rate.summary.df$easier, inversion.rate.summary.df$harder
)

inversion.rate.summary.df <- inversion.rate.summary.df %>% 
    rename(`Easier Course` = easier, `Harder Course` = harder) %>% 
    mutate(`Number of Comparisons` = inversions[1, ],
           `Times Harder was Shorter` = inversions[2, ],
           `Inversion Rate` = sprintf("%.1f%%", 100 * inversions[2, ] / inversions[1, ]))

knitr::kable(inversion.rate.summary.df)
```

More often than not, the Brown course is shorter than the Orange course. Additionally, the Yellow course is shorter than the White course about 5% of the time. This is slightly surprising, but not shocking: the primary differentiation between White and Yellow is ease of navigation. There are no recorded instances of Yellow being longer than Orange. Furthermore, we see that Brown is always shorter than Green, which in turn is always shorter than Red, which again in turn is always shorter than Blue. This makes sense: as noted earlier, the primary differentiator between advanced courses is distance, so the idea of a Brown longer than a Green or a Red longer than a Blue is nonsensical.

Finally, we see a few instances of unusual distance inversions across two course difficulty levels: twice when Yellow was longer than Brown (2015-04-12 at Lake Needwood, 3.19km Yellow vs 3.09km Brown; 2017-02-26 at Patapsco - Avalon, 4.27km Yellow vs 3.87km Brown), and once when Orange was longer than Green (2014-03-16 at Prince William Forest, 4.92km Orange vs 4.83km Green)!

## Shortest and Longest Courses

For fun, we can take a look at the shortest and longest courses of each color. The table below shows the shortest course for each color:

```{r shortest_course, echo=FALSE}
shortest.course.df <- classic.leg.summary.df %>% 
    group_by(course_clean) %>% 
    arrange(length) %>% 
    slice(1) %>% 
    ungroup() %>% 
    mutate(length = length / 1000) %>% 
    select(Color=course_clean, `Event Date`=event_date, Venue=location, `Length (km)`=length, Climb=climb,
           `Control Points`=control_points, `Course Setter`=course_setter)

knitr::kable(shortest.course.df)
```

The table below shows the longest course for each color:

```{r longest_course, echo=FALSE}
longest.course.df <- classic.leg.summary.df %>% 
    group_by(course_clean) %>% 
    arrange(desc(length)) %>% 
    slice(1) %>% 
    ungroup() %>% 
    mutate(length = length / 1000) %>% 
    select(Color=course_clean, `Event Date`=event_date, Venue=location, `Length (km)`=length, Climb=climb,
           `Control Points`=control_points, `Course Setter`=course_setter)

knitr::kable(longest.course.df)
```

# Course Slope

## Slope Description

Besides length, another salient characteristic of a course is how hilly it is. Typically, this is expressed as "climb", an estimate of the number of meters of elevation gain a runner will face on a near-optimal course route. To some extent, course distance is a bit of a fiction. The distance quoted for a course is the straight-line distance between each of the control points on the course, or the minimum possible distance that must be covered. In reality, the actual distance covered by any competitor is higher, perhaps significantly so: the actual route that a competitor takes from one control point to another will likely deviate for the sake of faster terrain, easier navigation, or both. In spite of these known limitations, course distance is still a useful metric.

Course climb is even more problematic than course climb. The traditional method for computing course climb is for the course setter to lay out a piece of string on a course map along (what is in their view) the optimal route, count the number of contour lines crossed by the string in an uphill direction, and sum them for an estimate of course distance.  This process involves significantly more subjectivity and room for error than distance measurements. However, despite its acknowledged imperfections, this measure of climb is better than any other readily available alternative, and is still useful for assessing courses.

Climb for each course is typically reported to the nearest 5 meters. Unsurprisingly, longer courses tend to have more climb associated with them, by virtue of the fact that they cover more distance. We can normalize for course distance by computing course slope: the ratio of climb (in meters) to course length (in meters). We must remember that slope represents meters of climb per meter of horizontal distance, and completely disregards vertical distance descended. Given that most orienteering courses have start and finish very close to each other, we must multiply quoted slope values by 2 to get the average terrain slope. For example, a 2km course with a 1km climb at a 10% grade followed by a 1km descent at a 10% grade would have 100m of climb over 2km of horizonal distance for a quoted slope of 0.05.

The plot below shows the distribution of course slope by course color. Analogously to the length plot shown previously, the dot shows the mean course slope, the heavy line shows the range from the 25th percentile to 75th percentile, and the lighter line shows the range from the 5th to the 95th percentile.

```{r course_slope_plot, echo=FALSE}
course.slope.summary.df <- classic.leg.summary.df %>% 
    mutate(slope = climb / length) %>% 
    group_by(course_clean) %>% 
    summarize(mu = mean(slope),
              q05 = quantile(slope, 0.05),
              q25 = quantile(slope, 0.25),
              q75 = quantile(slope, 0.75),
              q95 = quantile(slope, 0.95))

ggplot(course.slope.summary.df) +
    aes(x=course_clean) +
    geom_hline(yintercept=0) +
    geom_linerange(aes(ymin=q05, ymax=q95), alpha=0.7) +
    geom_linerange(aes(ymin=q25, ymax=q75), alpha=0.9, size=2) +
    geom_point(aes(y=mu), color="red",  size=1.5) +
    coord_cartesian(ylim=c(0, 0.06)) +
    labs(x="Course Color", y="Course Slope") +
    theme_bw()
```

We can see that mean course slope gradually increases over the recreational courses (from White to Orange), then levels off or slightly declines for the advanced courses (Brown to Blue).

## Flattest and Steepest Courses

Just as we did with the shortest and longest courses, we can for fun look for the flattest and steepest courses by color in our data.

The flattest courses by color are shown in the table below.

```{r flattest_course, echo=FALSE}
flattest.course.df <- classic.leg.summary.df %>% 
    mutate(slope = climb / length) %>% 
    group_by(course_clean) %>% 
    arrange(slope) %>% 
    slice(1) %>% 
    ungroup() %>% 
    mutate(length = length / 1000,
           slope = round(slope, 4)) %>% 
    select(Color=course_clean, `Event Date`=event_date, Venue=location, Slope=slope,
           Climb=climb, `Length (km)`=length, `Course Setter`=course_setter)

knitr::kable(flattest.course.df)
```

It comes as little surprise that five of the seven flattest courses are at Mason Neck, a notoriously flat venues. The only other venues on the list are Gunston Hall and Manassas, both of which have large, flat, open fields that recreational courses often use heavily.

The plot below shows the steepest course by color in the data:

```{r steepest_course, echo=FALSE}
steepest.course.df <- classic.leg.summary.df %>% 
    mutate(slope = climb / length) %>% 
    group_by(course_clean) %>% 
    arrange(desc(slope)) %>% 
    slice(1) %>% 
    ungroup() %>% 
    mutate(length = length / 1000,
           slope = round(slope, 4)) %>% 
    select(Color=course_clean, `Event Date`=event_date, Venue=location, Slope=slope,
           Climb=climb, `Length (km)`=length, `Course Setter`=course_setter)

knitr::kable(steepest.course.df)
```

We see here that unlike mean course slope, which tends to increase as a function of color, superlative course slope decreases as a function of course color. In other words, the steepest course on record is Yellow, not Brown or Red. This may be contrary to expectation, but there is a good explanation for this: we expect to see more variation in course slope for shorter courses than longer courses. For any given venue, some portions of the park are steeper than others. A short course may happen to be concentrated in the steeper or flatter portions of the park, and therefore have a more extreme slope rating. By contrast, longer courses must cover a greater portion of the park, and therefore necessarily be closer to the average slope for the park as a whole. Therefore, shorter courses are more likely to be exceptionally steep or exceptionally flat.

As well, there is more variety in terms of the venues with the steepest courses: four venues appear on the list, with no venue taking more than three spots.

Incidentally, we could also look for the courses with the most or least total climb, but as it turns out, such lists are very similar to the steepest and flattest lists shown.
